---
title: "Intro to R Workshop"
author: "TREC Training 2026"
date: 2026-03-18
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: show
    theme: cosmo
execute:
  echo: true
  warning: false
  message: false
editor: visual
fontsize: 11pt
---

```{r}
#| label: setup
source("training_source.R")  # loads tidyverse, rio, janitor, stringr, lubridate, scales, tidycensus, tidytext
```

## R Data Types and Structures

Types = kinds of values R can store (numeric, character, logical, factor).\
Structures = ways values are organized (vectors, lists, data frames).

### Numeric values

```{r}
1
3.14
-8
typeof(1)
```

### Character values

```{r}
"hello"
"30 minutes"
typeof("hello")
```

### Logical values

```{r}
TRUE
FALSE
typeof(TRUE)
```

### Vectors (single type)

```{r}
c(1, 2, 3)
c("a", "b", "c")
c(TRUE, FALSE)
c(1, "two", 3)   # coerced to character
```

### Factors (categorical values)

```{r}
# simple factor
modes <- factor(c("car","bike","car","bus"))
modes
levels(modes)

# ordered factor
sizes <- factor(c("small","large","medium"),
                levels = c("small","medium","large"),
                ordered = TRUE)
sizes
```

### Lists (can hold mixed types)

```{r}
my_list <- list(numbers = c(1,2,3), words = c("a","b"), mixed = list(TRUE, 5))
str(my_list)
```

### Data frames

```{r}
df <- data.frame(id = 1:3,
                 mode = c("walk","car","bus"),
                 distance = c(1.2, 5.0, 3.4))
df
str(df)
```

## Importing and Viewing Data

Import = load data from a file.\
Cleaning = prepare column names.\
Viewing = inspect rows and column types.

### Import with rio

```{r}
csv_path <- "../data/raw/presurvey_20260317.csv"
survey_raw <- rio::import(csv_path)
```

### Clean names and rename

```{r}
survey <- survey_raw |>
  janitor::clean_names() |>
  dplyr::rename(
    os = what_operating_system_are_you_using_for_this_workshop,
    prior_exp = which_of_the_following_are_true_regarding_your_experiences_with_r_prior_to_this_workshop,
    r_hope = what_do_you_hope_to_use_r_for_in_the_future,
    travel_mode = the_workshop_will_be_held_on_the_psu_campus_what_is_your_intended_mode_of_travel_to_the_workshop_check_all_that_apply,
    travel_direct = do_you_intend_to_travel_directly_from_home_to_the_workshop,
    distance = how_far_will_you_likely_travel,
    commute_time = what_is_the_approximate_travel_time_for_your_morning_commute_to_psu,
    drink = what_is_your_morning_drink_of_choice,
    movie_quote = what_is_the_first_movie_quote_that_comes_to_mind,
    pineapple_ok = pineapple_is_an_acceptable_pizza_topping
  )
```

### View

```{r}
head(survey)
glimpse(survey)
```

### Select and filter

```{r}
srv_small <- survey |>
  dplyr::select(os, prior_exp, r_hope, travel_mode, distance, commute_time, drink, pineapple_ok)

srv_small |>
  dplyr::filter(os == "Windows") |>
  head()
```

## Transforming Values

String extraction = pull digits or patterns from text.\
Conversion = turn extracted text into numbers.\
Text analysis (intro) = tokenize words and count with tidytext and stringr.

### Extract digits (stringr)

```{r}
survey <- survey |>
  dplyr::mutate(
    commute_digits  = stringr::str_extract(commute_time, "[0-9]+"),
    distance_digits = stringr::str_extract(distance,     "[0-9]+")
  )
```

### Convert written numbers (optional)

```{r}
# library(doseminer)
# survey <- survey |>
#   mutate(
#     commute_words_num  = doseminer::replace_numbers(commute_time),
#     distance_words_num = doseminer::replace_numbers(distance)
#   )
```

### Ensure optional columns exist (safe if optional step skipped)

```{r}
if (!"commute_words_num" %in% names(survey))  survey <- dplyr::mutate(survey,  commute_words_num  = NA_character_)
if (!"distance_words_num" %in% names(survey)) survey <- dplyr::mutate(survey, distance_words_num = NA_character_)
```

### Coerce to numeric

```{r}
survey <- survey |>
  dplyr::mutate(
    commute_min = as.numeric(dplyr::coalesce(commute_digits,  commute_words_num)),
    distance_mi = as.numeric(dplyr::coalesce(distance_digits, distance_words_num))
  )
```

### Factors (practical)

```{r}
# collapse messy OS strings to a few levels
ios_fct <- survey$os |>
  forcats::fct_collapse(Windows = c("Windows"),
                        macOS   = c("macOS", "Mac", "OS X"),
                        Linux   = c("Linux","Ubuntu","Debian","Fedora"),
                        Other   = c("Other","Unknown",""))
levels(ios_fct)
```

### Tidy text basics (tidytext + stringr)

```{r}
# tidytext is loaded from training_source.R
# Tokens = words from r_hope; lowercase; remove stop words
hope_tokens <- survey |>
  dplyr::select(r_hope) |>
  dplyr::mutate(r_hope = tolower(coalesce(r_hope, ""))) |>
  tidytext::unnest_tokens(word, r_hope) |>
  dplyr::anti_join(tidytext::stop_words, by = c("word" = "word")) |>
  dplyr::filter(word != "")

# Top words
hope_top <- hope_tokens |>
  dplyr::count(word, sort = TRUE)
head(hope_top, 15)
```

### Simple stringr counts

```{r}
# quick counts of mentions
table_map  <- stringr::str_detect(coalesce(survey$r_hope, ""), regex("map|mapping|gis", ignore_case = TRUE))
table_auto <- stringr::str_detect(coalesce(survey$r_hope, ""), regex("automate|automation|automating|automated", ignore_case = TRUE))

c(maps = sum(table_map), automation = sum(table_auto))
```

## Optional basics

Basics = small, useful operations

### Frequency tables

```{r}
survey |>
  dplyr::count(os, sort = TRUE)

survey |>
  dplyr::count(drink, sort = TRUE)
```

### Uniqueness

```{r}
dplyr::n_distinct(survey$os)

survey |>
  dplyr::distinct(os) |>
  head()
```

### Sampling rows

```{r}
dplyr::slice_sample(survey, n = 5)
```
